Tutorial for Runing Mouse Bone Marrow Dataset

### 2023-10-18 Demo_marrow_facs.Rmd
benchmark&figure_code

# Install & Load Packages

```{r}
# Install necessary tools if not already present
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
# Define package lists
bioc_packages <- c("Seurat",
                   "clusterProfiler", "AnnotationDbi", "ReactomePA", "org.Mm.eg.db")
github_packages <- c("satijalab/seurat-wrappers")
# Note: repeating packages from LMD_function.R as knitr does not install them properly
cran_packages <- c("igraph","ggplot2", "cowplot", "RColorBrewer",
                   "data.table", "dplyr", "patchwork",
                   "pheatmap", "ggplotify", "ggraph",
                   "ClusterR", "Rcpp", "RcppArmadillo", "tictoc")

# Install packages
sapply(cran_packages, function(pkg) if(!requireNamespace(pkg, quietly = TRUE)){install.packages(pkg)})
sapply(bioc_packages, function(pkg) if (!requireNamespace(pkg, quietly = TRUE)) BiocManager::install(pkg))
sapply(github_packages, function(pkg) if (!requireNamespace(strsplit(pkg, "/")[[1]][2], quietly = TRUE)) remotes::install_github(pkg))

@@

# Preparing input data

```{r}
if (!file.exists('facs_Marrow_seurat_tiss.Robj')) {
   # If load url doesn't work, please download it and load from your working directory
  options(timeout=6000) 
  download.file("https://figshare.com/ndownloader/files/13092380", 
                destfile = 'facs_Marrow_seurat_tiss.Robj', method = 'libcurl')
 }
load('facs_Marrow_seurat_tiss.Robj')
tiss <- UpdateSeuratObject(tiss)
```
We obtain the following objects from this Seurat Obj:
* `feature_space`: The output of PCA. This matrix contains the first 20 principal components (PCs).
* `visual_space`: a data.frame with t-SNE coordinates (2D) based on the first 20 PCs.
* `dat`: a matrix object with the log-normalized expression of genes (rows) in each cell (columns)
* `cell_label`: Any metadata you wish to visualize and examine related to cells (here we take the cell type annotation)
```{r}
DefaultAssay(tiss) <- "RNA"
n_dim = dim(tiss@reductions$pca@cell.embeddings)[2]
feature_space = as.matrix(tiss@reductions$pca@cell.embeddings[,1:n_dim])
visual_space = data.frame(tiss@reductions$tsne@cell.embeddings)
dat = as.matrix(tiss[[DefaultAssay(tiss)]]@data)
cell_label = tiss$cell_ontology_class
```

We process the `dat` by only keep the genes detected in more than 10 cells and less than 50% of cells (genes is "detected" in a cell only if its expression level is greater than the median expression level of that cell)

```{r}
Gene_detected_count <- apply(dat > apply(dat,2,median),1,sum)
selected_genes = (Gene_detected_count >= 10) & (Gene_detected_count <= ncol(dat) * 0.5)
dat = dat[selected_genes,,drop = FALSE]
```

# Running LMD Step by Step

This provides a step-by-step tutorial for a better understanding of the algorithm. If you want to generate results directly, please skip to 'Running LMD in One Step'

```{r}
source("./LMD_function.R", echo = F)
```

## Step1: Build KNN Graph

```{r}
# Construct knn graph
res = Symmetric_KNN_graph(knn = 5, feature_space = feature_space)

@@

## Step2: Diffuse Genes on the cell graph

```{r}
# Generate $P^{t}, t = 2,4,8,...,max_time$
P_ls = Obtain_Pls(W = W, max_time = 2^10)
# Generate initial gene distribution
rho = Rowwise_normalize(dat[,colnames(W)])
# Check One Gene Diffusion on Cell graph by time
gene = "Fcnb"
print(FeaturePlot_diffusion(coord = visual_space,init_state = rho[gene,],P_ls = P_ls,check_time = c(2,16,128),gene_name = gene))
```

## Step3: Obtain Diffusion KL score & Cumulative score

```{r}
# Obtain diffusion KL score
res = fast_calculate_multi_score(W = W, init_state = rho, P_ls = P_ls)

@@

# Running LMD in one step

```{r}
res = LMD(dat,feature_space)
```

# Identify Gene Modules

## Step1: Deriving Gene Modules

```{r}
# Select top1000 genes as local genes
local_gene = show_result_lmd(res, n = 1000)
local_gene = rownames(local_gene)

@@

# Re-name gene modules based on the order of htree
gene_partition = as.factor(gene_partition)
module_order = order(sapply(levels(gene_partition), function(mod) {
  median(which(gene_partition[gene_hree$order] == mod))
}))
levels(gene_partition)[module_order] = seq(nlevels(gene_partition))
gene_partition = factor(gene_partition,levels = seq(nlevels(gene_partition)))
rm(module_order)
```

### Visualize Heatmap

```{r}
p = pheatmap(as.matrix(dist),
              annotation_col = data.frame(Module = gene_partition), 
              cluster_rows = gene_hree, 
              cluster_cols = gene_hree,
              ann_colors = list(Module = setNames(colorRampPalette(brewer.pal(12, "Paired"))(nlevels(gene_partition)),levels(gene_partition))),
              treeheight_row = 0,
              col=colorRampPalette(c("red","white","blue4"))(99),
              show_colnames = FALSE, 
              show_rownames = FALSE,
              main = paste(distance_method,
                           clustering_method,sep = "-"),
              silent = TRUE )
as.ggplot(p)
```

## Step2: Computing Per-Cell Module Activity Scores

```{r}
knn = 10; major_vote = 5
A = Symmetric_KNN_graph(knn = knn, feature_space = feature_space)[[2]]
cell_block_loop = lapply(1:100,function(loop){
  GMM_subsampling(seed = loop, gene_partition, expr_dat = dat, cell_kNN_graph = A, major_vote)
})
cell_block_prop = Reduce(`+`, cell_block_loop) / length(cell_block_loop)
cell_block = cell_block_prop
```

### Visualize Module on Cell embeddings

```{r}
pl = lapply(levels(gene_partition),function(module){
  FeaturePlot_custom(value = cell_block[,module], coord = visual_space, value_name = "Module Score",title_name = sprintf("Module %s (%d genes)",module, sum(gene_partition == module)))
})
plot_grid(plotlist = pl,ncol = 10) # may takes long time to plot
```

## Step3: Identify Celltype-specific activated Modules

```{r}
# Create one-hot vector of Cell Type
one_hot_matrix <- sapply(unique(cell_label), function(x) as.integer(cell_label == x))

@@

# Calculate Jaccard index between module activity & cell type
jaccard_index <- apply(cell_block, 2, function(module_col) {
  apply(one_hot_matrix, 2, function(one_hot_col) {
    1 - philentropy::distance(rbind(module_col,one_hot_col), method = "jaccard", mute.message=TRUE)
  })
})
```

### Visualize Heatmap of Jaccard Index Matrix

```{r}
pheatmap(jaccard_index[rev(ordered_cell_types),], 
         cluster_rows=FALSE, cluster_cols=FALSE, 
         scale = "none", 
         color = colorRampPalette(rev(c("firebrick3", "white", "navy")))(100))
```

### Visualize How CellType-Module is Matched on Cell Embeddings

For each cell type, select the most specific modules for visualization (maximum jaccard index)

```{r}
tiss@meta.data = cbind(tiss@meta.data,one_hot_matrix)
jaccard_thred = 0.4

@@

pl <- lapply(rownames(jaccard_index)[rev(order(apply(jaccard_index,1,max)))], function(celltype){
  if(max(jaccard_index[celltype,]) > jaccard_thred){
    p1 = pl_ct[[celltype]]
    module_name = colnames(jaccard_index)[which.max(jaccard_index[celltype,])]
    p2 = pl_module[[module_name]]
    p = as.ggplot(p1) + as.ggplot(p2) + plot_annotation(paste0("Jaccard Index: ",round(jaccard_index[celltype,module_name],2)))
    p
  }
})
```

### Annotate Each Celltype-specific activated Modules with its corresponding Celltypes

```{r}
module_celltype = apply(jaccard_index,2,function(x){paste(names(which(x > 0.4)),collapse = ",")})
```

## Step4: Module Enrichment Analysis

### GO enrichment Analysis

```{r}
library("clusterProfiler")
library("ReactomePA")
library("org.Mm.eg.db")
library("AnnotationDbi")

@@

### Reactome enrichment Analysis

```{r}
library("ReactomePA")
library("org.Mm.eg.db")

@@

## Step5: Generate Gene Module Annotation Table

```{r}
df_module = data.frame(Module = levels(gene_partition))
df_module[,"Gene"] = unlist(lapply(levels(gene_partition),function(x){paste(names(gene_partition)[gene_partition == x],collapse = ",")}))
df_module[,"celltype"] = module_celltype[levels(gene_partition)]
df_module[top_go_clusters,"GO_description"] = unlist(lapply(top_enrich_go,function(x) paste(x[['Description']],collapse = ",")))
df_module[top_pathway_clusters,"Pathway_description"] = unlist(lapply(top_enrich_pathway,function(x) paste(x[['Description']],collapse = ",")))
df_module[is.na(df_module)] = ""
write.csv(df_module,file = "module_description.csv",row.names = FALSE)
```

>>>>

# Install & Load Packages
```{r}
# Install necessary tools if not already present
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
# Define package lists
bioc_packages <- c("Seurat")
github_packages <- c("satijalab/SeuratWrappers")

# Install packages
sapply(bioc_packages, function(pkg) if (!requireNamespace(pkg, quietly = TRUE)) BiocManager::install(pkg))
sapply(github_packages, function(pkg) if (!requireNamespace(strsplit(pkg, "/")[[1]][2], quietly = TRUE)) remotes::install_github(pkg))

@@

# Preparing input data
```{r}
# If load url doesn't work, please download it and load from your working directory
load(url("https://figshare.com/ndownloader/files/13092380"))
tiss <- UpdateSeuratObject(tiss)
```
We obtain the following objects from this Seurat Obj:
* `feature_space`: The output of PCA. This matrix contains the first 20 pricipal components (PCs).
* `visual_space`: a data.frame with t-SNE coordinates (2D) based on the first 20 PCs.
* `dat`: a matrix object with the log-normalized expression of genes (rows) in each cell (columns)
* `cell_label`: Any metadata you wish to visualize and examine related to cells (here we take the cell type annotation)
```{r}
DefaultAssay(tiss) <- "RNA"
n_dim = dim(tiss@reductions$pca@cell.embeddings)[2]
feature_space = as.matrix(tiss@reductions$pca@cell.embeddings[,1:n_dim])
visual_space = data.frame(tiss@reductions$tsne@cell.embeddings)
dat = as.matrix(tiss[[DefaultAssay(tiss)]]@data)
cell_label = tiss$cell_ontology_class
```
We process the `dat` by only keep the genes detected in more than 10 cells and less than 50% of cells (genes is "detected" in a cell only if its expression level is greater than the median expression level of that cell)
```{r}
Gene_detected_count <- apply(dat > apply(dat,2,median),1,sum)
selected_genes = (Gene_detected_count >= 10) & (Gene_detected_count <= ncol(dat) * 0.5)
dat = dat[selected_genes,,drop = FALSE]
```

# Running LMD Step by Step
This provides a step-by-step tutorial for a better understanding of the algorithm. If you want to generate results directly, please skip to 'Running LMD in One Step'
```{r}
source("./LMD_function.R", echo = F)
```
## Step1: Build KNN Graph
```{r}
# Construct knn graph
res = Symmetric_KNN_graph(knn = 5, feature_space = feature_space)

@@

## Step2: Diffuse Genes on the cell graph
```{r}
# Generate $P^{t}, t = 2,4,8,...,max_time$
P_ls = Obtain_Pls(W = W, max_time = 2^10)
# Generate initial gene distribution
rho = Rowwise_normalize(dat[,colnames(W)])
# Check One Gene Diffusion on Cell graph by time
gene = "Fcnb"
print(FeaturePlot_diffusion(coord = visual_space,init_state = rho[gene,],P_ls = P_ls,check_time = c(2,16,128),gene_name = gene))
```

## Step3: Obtain Diffusion KL score & Cumulative score
```{r}
# Obtain diffusion KL score
res = fast_calculate_multi_score(W = W, init_state = rho, P_ls = P_ls)

@@

# Running LMD in one step
```{r}
res = LMD(dat,feature_space)
```

# Identify Gene Modules
## Step1: Deriving Gene Modules
```{r}
# Select top1000 genes as local genes
local_gene = show_result_lmd(res, n = 1000)
local_gene = rownames(local_gene)

@@

# Re-name gene modules based on the order of htree
gene_partition = as.factor(gene_partition)
module_order = order(sapply(levels(gene_partition), function(mod) {
  median(which(gene_partition[gene_hree$order] == mod))
}))
levels(gene_partition)[module_order] = seq(nlevels(gene_partition))
gene_partition = factor(gene_partition,levels = seq(nlevels(gene_partition)))
rm(module_order)
```
### Visualize Heatmap
```{r}
p = pheatmap(as.matrix(dist),
              annotation_col = data.frame(Module = gene_partition), 
              cluster_rows = gene_hree, 
              cluster_cols = gene_hree,
              ann_colors = list(Module = setNames(colorRampPalette(brewer.pal(12, "Paired"))(nlevels(gene_partition)),levels(gene_partition))),
              treeheight_row = 0,
              col=colorRampPalette(c("red","white","blue4"))(99),
              show_colnames = FALSE, 
              show_rownames = FALSE,
              main = paste(distance_method,
                           clustering_method,sep = "-"),
              silent = TRUE )
as.ggplot(p)
```

## Step2: Computing Per-Cell Module Activity Scores
```{r}
knn = 10; major_vote = 5
A = Symmetric_KNN_graph(knn = knn, feature_space = feature_space)$'adj_matrix'
cell_block_loop = lapply(1:100,function(loop){
  cat("\n",loop,"\n")
  GMM_subsampling(seed = loop, gene_partition, expr_dat = dat, cell_kNN_graph = A, major_vote)
})
cell_block_prop = Reduce(`+`, cell_block_loop) / length(cell_block_loop)
cell_block = cell_block_prop
```
### Visualize Module on Cell embeddings
```{r}
pl = lapply(levels(gene_partition),function(module){
  FeaturePlot_custom(value = cell_block[,module], coord = visual_space, value_name = "Module Score",title_name = sprintf("Module %s (%d genes)",module, sum(gene_partition == module)))
})
plot_grid(plotlist = pl,ncol = 10) # may takes long time to plot
```

## Step3: Identify Celltype-specific activated Modules
```{r}
# Create one-hot vector of Cell Type
one_hot_matrix <- sapply(unique(cell_label), function(x) as.integer(cell_label == x))

@@

# Calculate Jaccard index between module activity & cell type
jaccard_index <- apply(cell_block, 2, function(module_col) {
  apply(one_hot_matrix, 2, function(one_hot_col) {
    1 - philentropy::distance(rbind(module_col,one_hot_col), method = "jaccard")
  })
})
jaccard_index = jaccard_index[rev(ordered_cell_types),]
rm(ordered_cell_types)
```

### Visualize Heatmap of Jaccard Index Matrix
```{r}
pheatmap(jaccard_index, 
         cluster_rows=FALSE, cluster_cols=FALSE, 
         scale = "none", 
         color = colorRampPalette(rev(c("firebrick3", "white", "navy")))(100))
```

### Visualize How CellType-Module is Matched on Cell Embeddings
For each cell type, select the most specific modules for visualization (maximum jaccard index)
```{r}
tiss@meta.data = cbind(tiss@meta.data,one_hot_matrix)
jaccard_thred = 0.4

@@

pl <- lapply(rownames(jaccard_index)[rev(order(apply(jaccard_index,1,max)))], function(celltype){
  if(max(jaccard_index[celltype,]) > jaccard_thred){
    p1 = pl_ct[[celltype]]
    module_name = colnames(jaccard_index)[which.max(jaccard_index[celltype,])]
    p2 = pl_module[[module_name]]
    p = as.ggplot(p1) + as.ggplot(p2) + plot_annotation(paste0("Jaccard Index: ",round(jaccard_index[celltype,module_name],2)))
    p
  }
})
```
### Annotate Each Celltype-specific activated Modules with its corresponding Celltypes
```{r}
module_celltype = apply(jaccard_index,2,function(x){paste(names(which(x > 0.4)),collapse = ",")})
```

## Save Intermediate data
```{r}
save(gene_partition,dist,gene_hree,cell_block,jaccard_index,file = "./marrow_result/module_result.rdata")
```

## Step4: Module Enrichment Analysis
### GO enrichment Analysis
```{r}
if (!requireNamespace("clusterProfiler", quietly = TRUE)) BiocManager::install("clusterProfiler")
library(clusterProfiler)

@@

### Reactome enrichment Analysis
```{r}
if (!requireNamespace("ReactomePA", quietly = TRUE)) BiocManager::install("ReactomePA")
if (!requireNamespace("org.Mm.eg.db", quietly = TRUE)) BiocManager::install("org.Mm.eg.db")
library("clusterProfiler")
library("ReactomePA")
library("org.Mm.eg.db")

@@

## Step5: Generate Gene Module Annotation Table
```{r}
df_module = data.frame(Module = levels(gene_partition))
df_module[,"Gene"] = unlist(lapply(levels(gene_partition),function(x){paste(names(gene_partition)[gene_partition == x],collapse = ",")}))
df_module[,"celltype"] = module_celltype[levels(gene_partition)]
df_module[top_go_clusters,"GO_description"] = unlist(lapply(top_enrich_go,function(x) paste(x[['Description']],collapse = ",")))
df_module[top_pathway_clusters,"Pathway_description"] = unlist(lapply(top_enrich_pathway,function(x) paste(x[['Description']],collapse = ",")))
df_module[is.na(df_module)] = ""
write.csv(df_module,file = "module_description.csv",row.names = FALSE)
```
